FLM for growth as function of SPEI

### 2021-01-26 analysis/paper/vital rate models.Rmd
use all data for survival, but subset data for growth and flowering to be only plants that survived (and therefore have size measurements).

```{r}
demog_full <-
  demog_full %>% 
  filter(ht < 200, ht_prev < 200) %>% 
  filter(shts > 0, shts_prev > 0, ht > 0, ht_prev > 0) %>% 
  mutate(size = shts*ht, size_prev = shts_prev * ht_prev, log_size = log(size), log_size_prev = log(size_prev))
```

@@

# Speed up models with bam()

@@

# Growth

Start with no random effects

Using a scaled-t family distribution because residuals were leptokurtic with a gaussian.

@@

Fit the GAM with crossbasis smooth for lagged drought.

Preliminary modeling showed that if year is included as random effect it is concurve with SPEI history (which would mean biased p-values and coefficients).

```{r}
g_post_cf <-
  bam(log_size ~ 
        s(log_size_prev) + 
        # s(year_fac, bs = "re") + 
        s(plot, bs = "re") + #random effect of plot
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      # family = gaussian(link = "identity"),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_cf,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

@@

AIC for... 
Gaussian: 40909.27. looks leptokurtic
scaled t: 37284.47.  Waaay better

@@

```{r}
g_post_1ha <-
  bam(log_size ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog_post_1ha,
      select = TRUE,
      method = "REML",
      cluster = cl)
```

@@

newdata_covar_cf <- 
  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
         plot = "newplot",
         year_fac = "newyear", 
         spei_history = mean(demog_post_cf$spei_history),
         L = demog_post_cf$L[1:100, ])

out <- augment(g_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)

@@

out <- augment(g_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)

@@

# Survival

not significant!  I should try select=TRUE here.

Uh...There might be something wrong with how I've determined survival.  I should revisit this.

```{r}
out <- augment(s_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)

These are not significant, so probably don't need to actually make these plots.  I'm mostly just curious what a non-significant crossbasis function looks like.

@@

```{r}
f_post_cf <-
  bam(flwr ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") + 
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_cf,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

@@

```{r}
f_post_1ha <-
  bam(flwr ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog_post_1ha,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

>>>>

```{r}
demog_full <-
  demog_full %>% 
  filter(ht < 200 | is.na(ht), ht_prev < 200 | is.na(ht_prev)) %>%
  filter(
    # having no aboveground biomass is possible, but categorically different than just being small. Exclude plants with 0 shoots or 0 height
    shts_prev > 0,
    ht_prev > 0,
    #for survival data, must include plants with NA for shts and height (i.e. dead plants)
    shts > 0 | is.na(shts), 
    ht > 0 | is.na(ht)
  ) %>%
  mutate(size = shts*ht, size_prev = shts_prev * ht_prev, log_size = log(size), log_size_prev = log(size_prev))
```

@@

# Inspect raw data

```{r}
demog_post %>% ggplot(aes(x = surv)) + geom_bar() + facet_wrap(~habitat)
```
```{r}
demog_post %>% ggplot(aes(x = habitat, y = log_size)) + geom_boxplot()
```
```{r}
demog_post %>% ggplot(aes(x = flwr)) + geom_bar() + facet_wrap(~habitat)
```
# Speed up models with bam()

@@

# Survival
`surv` is 1 in `year`s when a plant is alive and 0 in the first `year` when the plant is dead.  These GAMs model the survival of a plant (over the past year) as a function of size in the previous census and SPEI values starting at february of the current year going back 36 months.
## Fit GAMs

newdata_covar_cf <- 
  tibble(log_size_prev = seq(sr[1], sr[2], length.out = 100),
         plot = "newplot",
         year_fac = "newyear", 
         spei_history = mean(demog_post_cf$spei_history),
         L = demog_post_cf$L[1:100, ])
```

out <- augment(s_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)

# Growth
For mature plants that are alive...

```{r}
demog2_post_cf <- demog_post_cf %>% filter(surv == 1, !is.na(log_size))
demog2_post_1ha <- demog_post_1ha %>% filter(surv == 1, !is.na(log_size))
```

@@

Fit the GAM with crossbasis smooth for lagged drought.
Using a scaled-t family distribution because residuals were leptokurtic with a gaussian.
Preliminary modeling showed that if year is included as random effect it is concurve with SPEI history (which would mean biased p-values and coefficients).

The transformed response and identity link is actually better than a log-link, according to Ellner, Childs and Rees.

```{r}
g_post_cf <-
  bam(log_size ~ 
        s(log_size_prev) + 
        # s(year_fac, bs = "re") + 
        s(plot, bs = "re") + #random effect of plot
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal-ish response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      # family = gaussian(link = "identity"),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog2_post_cf,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

@@

AIC for... 
Gaussian: ~40909.27. looks leptokurtic
scaled t: ~37284.47.  Waaay better

@@

```{r}
g_post_1ha <-
  bam(log_size ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #unimodal response to drought, but one knot per month lag.
          xt = list(bs = "cr")),
      family = scat(link = "identity"), #like leptokurtic gaussian.
      data = demog2_post_1ha,
      select = TRUE,
      method = "REML",
      cluster = cl)
```

@@

```{r}
out <- augment(g_post_cf, newdata = newdata_covar_cf, se_fit = TRUE)

@@

```{r}
out <- augment(g_post_1ha, newdata = newdata_covar_1ha, se_fit = TRUE)

@@

@@

```{r}
f_post_cf <-
  bam(flwr ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") + 
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog2_post_cf,
      method = "REML",
      select = TRUE,
      cluster = cl)
```

@@

```{r}
f_post_1ha <-
  bam(flwr ~ 
        s(log_size_prev) +
        # s(year_fac, bs = "re") +
        s(plot, bs = "re") +
        s(spei_history, L,
          bs = "cb",
          k = c(3, 35),  #cubic response (at most) to drought, one knot per month lag.
          xt = list(bs = "cr")),
      family = binomial,
      data = demog2_post_1ha,
      method = "REML",
      select = TRUE,
      cluster = cl)
```
